Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-17 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Working on issue

\IEEEPARstart{N}{owadays} there is a big and increasing number of
distributed evolutionary computation software frameworks available on
the market, usually as open source systems \cite{Parejo12Survey}. Most of these systems have 
two features: they are {\em desktop} applications, that is, they must
be compiled or run directly on the operating system layer and, second,
they assume that the nodes that are going to run the system are under
your control or at least you have access to them.

There is, then, space for systems that do not make any of these
assumptions, such as the one presented in this paper, which will be
provisionally named {\sf NodIO}. {\sf NodIO} is first a two tier client/server 
application, in which clients can possibly run in a browser. 
Therefore, it can be run in any web-enabled device, including many over which 
the scientist has no control.

@@

%PABLO: He añadido el siguiente párrafo para el issue #31
There exist other volunteer frameworks written in compiled languages,
such as C++ or Java, that can achieve more performance than
JavaScript. However, one of the advantages of our system is that
clients do not need to install anything in their machines, they just
need to open a regular web page  
% Somewhat redundant 
% Javascript is the only language that can be executed in this way without any installation, 
% also removed the default, redundant with next sentence
in most browsers, such as Chrome, Safari or Firefox, that are
available in almost every desktop or mobile device. Other advantage is
that the source code can be directly examined by the volunteer, not
being compiled as other options (such as Java Applets, Flash scripts or
ActiveX). Also, as we discuss in this paper, it can provide some extra
features such as enhanced security or operating system independence.

%PABLO: hay que enlazar con el siguiente párrafo
Those desktop clients have also the same issues as any kind of
volunteer computing system, including our own:  these issues are
related to security as well as
safety, which introduces a {\em social} aspect to the design of a
computing system that makes it a {\em techno-social system}
\cite{vespignani2009predicting}. In this paper our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page that runs the system
determine its performance and how it can be modeled and, eventually,
predicted.

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using Java
compiled programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use and
reasonable performance. Of all these, probably {\em hierarchy}, that
is, the capability of using preferentially resources in the same
organization could be either disregarded or included in adaptivity,
leaving us with eight desirable properties.

The fulfillment of these properties could be achieved using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first, but embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed using desktop
applications. Baratloo et al. tried in 1996 to create a system with
the aforementioned qualities called {\em Charlotte}
\cite{baratloo1996charlotte}. This system, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing systems PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system and access to a common data store in the
server. Since the Java virtual machine is a sandbox, it is safe to the user
and running a compiled bytecode guarantees a certain level of
performance.
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. This
paper was later extended and published in a journal in 1999
\cite{baratloo1999charlotte}, but it is interesting to note that
already in the early stages of the web the possibility of
browser-based distributed computing was already researched.

@@

Recent works propose the use of cloud computing services for the components of
client-server distributed EAs. Cloud resources
can reduce operational costs by outsourcing both hardware and software maintenance
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what
is available in most laboratories, allowing researches to
scale algorithms at reasonable costs. Researchers have proposed the use of
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards}
and DropBox~\cite{mericloud}. The use of JavaScript to implement
server side components
gives software developers certain advantages in the cloud
paradigm, since it is a first-class citizen
in most Platforms as a Service (PaaS) products as the {\tt Node.js} framework is supported by most. \cite{wood13:nodejs:paas}. Many web services
have native JavaScript libraries too, and also use JSON (JavaScript Object Notation) for data interchange. Another advantage is that
developers can use the same software tools for both client and server
side development.

@@

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}, which is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or farming out evaluations to all
the nodes available. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which can, of course, be studied in
a more general distributed computing context
\cite{nogueras2015studying} or including it in a more general study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

On the other hand, measuring performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. But the raw material of
volunteer computing, number of users and the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at larger scale in both quantity of volunteers and scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
system itself.

As far as we know, this paper presents one of the few experiments that
uses computational resources that are as dissimilar as smartphones and
powerful laptops 
or desktop computers in a research center. The methodology used to
gather resources and the algorithms used will be described next, as
well as the results obtained with the setup.

@@

In general, a volunteer-based distributed evolutionary computation
system based on the browser is simply a client-server architecture
whose client is, or can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; that is
why we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M}, resulting that, in fact, JavaScript can be
faster than compiled languages such as Scala and, in any case, provide
a performance that is comparable to other languages employed usually
in evolutionary computation and numerical optimization such as Python. Compiled languages such as
Java or C might be, in fact, faster, but the fact that a single-user
performance is higher is more than offset by the fact that the number
of spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets, is several orders of magnitude
higher, which is the objective in this kind of systems. JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala or even Java, and them compile it to
JavaScript.

As a volunteer based system, there are some issues worth 
noting \cite{sarmenta2001volunteer,web:BOINC} :\begin{enumerate}
\item Volunteers are anonymous, only the IP address sent in 
each request is known.
\item As anonymous entities, volunteers are not accountable. 
If a volunteer misbehaves in some way the provider cannot 
prosecute or discipline the volunteer. In the case of JavaScript,
this issue is agravated because even if it is obfuscated, sorce code
and data is sent to clients. 
\item Participants must also trust the application provider, 
regarding invation of privacy, the integrity of their devices 
and intended application. 
\end{enumerate}

@@

In this sense, in this paper we propose the {\sf NodIO} framework, a
cloud or bare metal based volunteer evolutionary computing system
based on the {\sf NodEO} library, whose architecture has been
developed using JavaScript on the client as well as the server sides.
All parts of the system are free and available with a free license
from \url{https://github.com/JJ/splash-volunteer}.

@@

Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several {\em
  routes} 
  that can be called for storing and retrieving information (the CRUD cycle:
  create, request, update and delete) from the server. 
  A JSON data format is used for the communication between 
  clients and the server. There are two kinds of information:
  {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
  GET}ing a random chromosome from it, and {\em information} related
  to performance and state of the experiments. It also performs logging
  duties, but is basically a very lightweight and high performance
  data store \cite{jj:idc:lowcost}.
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found.
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with a random
  population and sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request).
\end{enumerate}

@@

We found that the JavaScript ecosystem is mature enough for developing non-trivial 
code, but developers have to consider the differences betwwen VMs implementations
and also use certain libraries in order to increase the precision and repeatability of
algorithms. As for the performance of the JavaScript implementaion against Java and Matlab
we conducted first the basic test described earlier in a 3.7 GHz Quad-Core Intel Xeon E5 processor running OS X 10.10.5, 
using Java(TM) SE Runtime Environment (build 1.8.0\_25-b17), 
MatLab version R2015a for Mac OSX, Node FrameWork version 0.12.2, Google Chrome Version 46.0.2490.86 (64-bit) 
and Firefox version 41.0.2. The performance of the time required for 10,000 FEs is presented in Figure \ref{fig:f15_times}.
In our test the Matlab implementation had the best performance with an average of 935ms, followed by Java with
991ms, JavaScipt took 32\% more time than Java with 1240ms in Chrome using a Web Worker and 1,270ms using Node.js. 
The results also show that there is not much difference between the Node.Js implemenation an that of a
browser. There is also not much difference between running the code in the main thread or as workers, but
when running the experiment in Firefox a dialog alerted of the script taking to long, showing that this
kind of processing is not intended for the main thread. Also there is not much diference when running two
Web Workers in parallel taking 1279ms each. This experiment shows that JavaScript is viable 
language for developing complex mathematical functions for running in the browser. 
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{f15_times.png}
\caption{ Runtime of 10,000 function evaluations (FEs) for $F_{15}$} 
\label{fig:f15_times}
\end{figure}

>>>>

\IEEEPARstart{N}{owadays} there is a big and increasing number of
distributed evolutionary computation software frameworks available on
the market, usually as open source systems \cite{Parejo12Survey}. Most of these systems have 
two features: they are {\em desktop} applications, that is, they must
be compiled or run directly on the operating system layer and, second,
they assume that the nodes that are going to run the system are under
the control of developers or they at least have access to them.

There is, then, space for systems that do not make any of these
assumptions, such as the one presented in this paper, which will be
provisionally named {\sf NodIO}. {\sf NodIO} is first a two tier client/server 
application, in which clients can run in a browser. 
Therefore, they can embedded any web-enabled device, including many over which 
the developers have no control.

@@

%PABLO: He añadido el siguiente párrafo para el issue #31
There exist other volunteer frameworks written in compiled languages,
such as C++ or Java, that can achieve better performance than
JavaScript. However, one of the advantages of our system is that
clients do not need to install anything in their machines, they just
need to open a web page  
% Somewhat redundant 
% Javascript is the only language that can be executed in this way without any installation, 
% also removed the default, redundant with next sentence
in a compatible browser, such as Chrome, Safari or Firefox, which are
available in almost every desktop or mobile device. Other advantage is
that the source code can be directly examined by volunteers, because is not
compiled as other options (such as Java Applets, Flash scripts or
ActiveX). Also, as we discuss in this paper, it can provide some extra
features such as enhanced security or operating system independence.

%PABLO: hay que enlazar con el siguiente párrafo
Desktop clients have the same issues as any 
volunteer computing system, including our own: these issues are
related to security as well as
safety, which introduce a {\em social} aspect to the design of a
computing system that makes it a {\em techno-social system}
\cite{vespignani2009predicting}. In this paper our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page that runs the system
determine its performance and how it can be modeled and, eventually,
predicted.

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use and
reasonable performance. Of all these, probably {\em hierarchy}, that
is, the capability of using preferentially resources in the same
organization could be either disregarded or included in adaptivity,
leaving us with eight desirable properties.

The fulfillment of these properties could be achieved using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first, but embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed by using desktop
applications. Baratloo et al. tried in 1996 to create a system with
the aforementioned qualities called {\em Charlotte}
\cite{baratloo1996charlotte}. This system, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing systems PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system and access to a common data store in the
server. Since the Java virtual machine is a sandbox, it is safe to the user
and as it runs compiled bytecode it also guarantees a certain level of
performance.
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. This
paper was later extended and published in a journal in 1999
\cite{baratloo1999charlotte}, but it is interesting to note that
already in the early stages of the web the possibility of
browser-based distributed computing was already researched.

@@

Recent works propose the use of cloud computing services for the components of
client-server distributed EAs. Cloud resources
can reduce operational costs by outsourcing both hardware and software maintenance
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what
is available in most laboratories, allowing researches to
scale algorithms at reasonable costs. Researchers have proposed the use of
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards}
and DropBox~\cite{mericloud}. The use of JavaScript to implement
server side components
gives software developers certain advantages in the cloud
paradigm, since it is a first-class citizen
in most Platforms as a Service (PaaS) products as the {\tt Node.js} framework is supported by most of them. \cite{wood13:nodejs:paas}. Many web services
have native JavaScript libraries too, and also use the JSON (JavaScript Object Notation) format for data interchange. Another advantage is that
developers can use the same software tools for both client and server
side development.

@@

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}, which is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or farming out evaluations to all
the nodes available. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or including it in a more general study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

On the other hand, measuring the performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. But the raw material of
volunteer computing, number of users and the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
system itself.

As far as we know, this paper presents one of the few experiments that
uses computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. The methodology used to
gather resources and the algorithms used will be described next, as
well as the results obtained with the initial setup.

@@

In general, a volunteer-based distributed evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; that is
why we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M}, resulting that, in fact, JavaScript can be
faster than compiled languages such as Scala and, in any case, provide
a performance that is comparable to other languages employed usually
in evolutionary computation and numerical optimization such as Python. Compiled languages such as
Java or C might be, in fact, faster, but the fact that a single-user
performance is higher is more than offset by the fact that the number
of spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets, is several orders of magnitude
higher, which is the objective in this kind of systems. JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala or even Java, and them compile it to
JavaScript. For completness, an experiment comparing the performance of
a JavaScript implementation versus Java and Matlab is included in section
\ref{sec:w2}.

As a volunteer based system, there are some issues worth 
noting \cite{sarmenta2001volunteer,web:BOINC} :\begin{enumerate}
\item Volunteers are anonymous, only the IP address sent in 
each request is known.
\item As anonymous entities, volunteers are not accountable. 
If a volunteer misbehaves in some way the provider cannot 
prosecute or discipline the volunteer. In the case of JavaScript,
this issue is aggravated because even if it is obfuscated, source code
and data is sent to clients. 
\item Participants must also trust the application provider, 
regarding invasion of privacy, the integrity of their devices 
and intended application. 
\end{enumerate}

@@

In this sense, in this paper we propose the {\sf NodIO} framework, a
cloud or bare metal based volunteer evolutionary computing system
based on the {\sf NodEO} library, whose architecture has been
developed using JavaScript on the client as well as the server.
All parts of the system are free and available with a free license
from \url{https://github.com/JJ/splash-volunteer}.

@@

Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several {\em
  routes} 
  that can be called for storing and retrieving information (the CRUD cycle:
  create, request, update and delete) from the server. 
  A JSON data format is used for the communication between 
  clients and the server. There are two kinds of information:
  {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
  GET}ing a random chromosome from it, and {\em information} related
  to the performance and state of the experiments. It also performs logging
  duties, but is basically a very lightweight and high performance
  data store \cite{jj:idc:lowcost}.
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found.
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with a random
  population and then sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request).
\end{enumerate}

@@

We found that the JavaScript ecosystem is mature enough for developing non-trivial 
code, but developers have to consider the differences betwwen VMs implementations
and also use certain libraries in order to increase the precision and repeatability of
algorithms. As for the performance of the JavaScript implementaion against Java and Matlab
we conducted first the basic test described earlier in a 3.7 GHz Quad-Core Intel Xeon E5 processor running OS X 10.10.5, 
using Java(TM) SE Runtime Environment (build 1.8.0\_25-b17), 
MatLab version R2015a for Mac OSX, Node FrameWork version 0.12.2, Google Chrome Version 46.0.2490.86 (64-bit) 
and Firefox version 41.0.2. The performance of the time required for 10,000 FEs is presented in Figure \ref{fig:f15_times}.
In our test the Matlab implementation had the best performance with an average of 935ms, followed by Java with
991ms, JavaScript took 32\% more time than Java with 1240ms in Chrome using a Web Worker and 1,270ms using Node.js. 
The results also show that there is not much difference between the Node.Js implementation an that of a
browser. There is also not much difference between running the code in the main thread or as workers, but
when running the experiment in Firefox a dialog alerted of the script taking to long, showing that this
kind of processing is not intended for the main thread. Also there is not much difference when running two
Web Workers in parallel taking 1279ms each. This experiment shows that JavaScript is viable 
language for developing complex mathematical functions for running in the browser. 
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{f15_times.png}
\caption{ Runtime of 10,000 function evaluations (FEs) for $F_{15}$} 
\label{fig:f15_times}
\end{figure}
